---
title: "2020 Election Visualizations"
author: "William Cichowski"
date: 2021-11-03T21:13:14-05:00
categories: ["R"]
tags: ["R Markdown", "SQL"]
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE)
```
<br>

## Project Outline

**In this project, we were given the datasets and asked to recreate these visuals. Many people were skeptical about the advent of mail-in ballots in the 2020 presidential election. Specifically, many pundits, news agencies, and politicians stated that mail-in ballots benifited democrats for whatever reason. In the following project, we take the state of Georgia, and figure out by county, how large the percentage of mail-in voters was.** 
<br>
```{r load-packages-data-seed, eval=TRUE, message=FALSE, warning=FALSE}
# Load package(s)
library(tidyverse)
library(janitor)
library(grid)
library(jpeg)
library(sf)
library(patchwork)
library(ggforce)

# Load datasets
ga_election_dat <-
read_csv(file = "ga_election_data.csv") %>% 
  clean_names() %>% 
  mutate(prop_of_early = (absentee_by_mail_votes + advanced_voting_votes)/ total_votes) %>% 
  select(-contains("_votes"))
```
<br>
```{r, fig.height=7, fig.width=7}
# loading data for Georgia counties
ga_counties <- tigris::counties(state = "GA", cb = TRUE, progress_bar = FALSE) %>% 
  clean_names()

# Biden info table & map

# left joining the two tables together with biden data
ga_biden_new <-
  ga_counties %>% 
  left_join(
    ga_election_dat %>% 
      filter(candidate == "Joseph R. Biden"),
    by = c("name" = "county"))

# creating the plot
biden_plot <- ga_biden_new %>% 
  ggplot() +
  geom_sf(aes(fill = prop_of_early)) +
  scale_fill_gradient2(low = "#1AFF1A",
                       high = "#5D3A9B",
                       midpoint = .75,
                       name = NULL,
                       limits = c(.5, 1),
                       breaks = c(.5, .75, 1),
                       labels = scales::percent_format(),
  ) +
  labs(
    title = "Joseph R. Biden",
    subtitle = "Democratic Nominee"
  ) +
  theme_void() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 12, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 12))

# left joining the two tables together with trump data
ga_trump_new <-
  ga_counties %>% 
  left_join(
    ga_election_dat %>% 
      filter(candidate == "Donald J. Trump"),
    by = c("name" = "county"))

# creating the plot
trump_plot <-
  ga_trump_new %>% 
  ggplot() +
  geom_sf(aes(fill = prop_of_early)) +
  scale_fill_gradient2(low = "#1AFF1A",
                       high = "#5D3A9B",
                       midpoint = .75,
                       name = NULL,
                       limits = c(.5, 1),
                       breaks = c(.5, .75, 1),
                       labels = scales::percent_format(),
  ) +
  labs(
    title = "Donald J. Trump",
    subtitle = "Republican Nominee"
  ) +
  theme_void() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 12, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 12))

comparision <- biden_plot + trump_plot
comparision + plot_annotation(
  title = 'Percentage of votes From early voting',
    theme = theme(plot.title = element_text(size = 24, face = 'bold')))
```

**The insight we have learned was that both nominees had a large portion of their votes from mail in ballots 
(over 50% in almost every county for both candidates). This makes the argument that the democrats used mail in 
ballots to illegally cast extra votes less believable.**


